---
title: "Session 1: Analysing serological data"
format: html
editor: visual
execute:
  warning: false
  message: false
  eval: false
---

# Introduction to serology

In this practical, we are going to explore the concept of the **force of infection** and how it relates to seroprevalence.

## Objectives

The aims of this practical are to:

- Load and analyse a serological dataset in R

- Understand how changing the force of infection affects the proportion seropositive

## Set-up: load packages and datasets

First, we will load in the R packages that we need

```{r}
library(tidyverse)
```

```{r}
sero_survey <- read.csv("../data/SeroStudy.csv")
head(sero_survey) # look at the first few rows of the dataset
```

1.  What is the range and mean age of participants?

::: {.callout-tip collapse="true"}
## Hint

Use `range()` or `min()` and `max()`
:::

```{r}
sero_survey |>
  summarise(
    min_age = min(Age),
    max_age = max(Age),
    mean_age = mean(Age)
  )
```

2.  What proportion of male participants are seropositive?

::: {.callout-tip collapse="true"}
## Hint

You can use `table()`
:::

```{r}
sero_survey |>
  filter(Sex == "Male") |>
  summarise(prop_seropositive = mean(IgG))
```

3.  What proportion of female participants are seropositive?

```{r}
sero_survey |>
  filter(Sex == "Female") |>
  summarise(prop_seropositive = mean(IgG))
```

4.  Calculate the proportion seropositive by single age

::: {.callout-tip collapse="true"}
## Hint

You can use a `for` loop
:::

```{r}
prop_by_age <- sero_survey |>
  group_by(Age) |>
  summarise(prop_seropositive = mean(IgG))

prop_by_age |>
  ggplot(aes(x = Age, y = prop_seropositive)) +
  geom_point()
```

5.  Using the following relationship, plot the expected proportion seropositive if there was a force of infection of 0.05 in a community:

$$
P(seropositive) = 1 - e^{(-\lambda \cdot age)}$$ \$\$

::: {.callout-tip collapse="true"}
## Hint

Create an object `lambda = 0.05` and use a vector of ages between 1 and 15 `ages <- 1:15`
:::

```{r}
lambda <- 0.05
ages <- 1:15

expected_seropos <- tibble(
  Age = ages,
  prop_seropositive = 1 - exp(-lambda * ages)
)

expected_seropos |>
  ggplot(aes(x = Age, y = prop_seropositive)) +
  geom_line()
```

6.  Plot the line from no. 5 on top of the observed proportion seropositive by age – how similar does it look?

```{r}
prop_by_age |>
  ggplot(aes(x = Age, y = prop_seropositive)) +
  geom_point() +
  geom_line(data = expected_seropos, colour = "red")
```

7.  Try a few different values for $$\lambda$$ and find one that works well

```{r}
lambdas <- seq(0.01, 0.15, by = 0.01)

expected_seropos_multi <- lambdas |>
  map_dfr(\(lambda) tibble(
    lambda = lambda,
    Age = ages,
    prop_seropositive = 1 - exp(-lambda * ages)
  ))

prop_by_age |>
  ggplot(aes(x = Age, y = prop_seropositive)) +
  geom_point() +
  geom_line(
    data = expected_seropos_multi,
    aes(group = lambda),
    colour = "red", alpha = 0.4
  )
```

## Stretch questions

8.  Using this value of $$\lambda$$ calculate the proportion of 30 year olds you would expect to be seropositive

```{r}
lambda <- 0.05 # replace with the value of lambda you found above

tibble(Age = 30) |>
  mutate(prop_seropositive = 1 - exp(-lambda * Age))
```

9.  Using this value of theta, calculate the probability of being infected for each age between 1 and 60

```{r}
ages_60 <- 1:60

prob_infected <- tibble(
  Age = ages_60,
  prob_infection = 1 - exp(-lambda * ages_60)
)

prob_infected
```

10. Assuming that there are 10,000 of each age between 1 and 60 (and no one older) – so 10,000 1 year olds, 10,000 2 year olds etc, calculate who many infections you see in total across all age groups in this population

```{r}
pop_uniform <- prob_infected |>
  mutate(n_people = 10000,
         n_infected = n_people * prob_infection)

pop_uniform |>
  summarise(total_infected = sum(n_infected))
```

11. Repeat the last question but assume this time there are 20,000 in each age between 1 and 10, 10,000 between 11-50 and no one older (ie a younger population of the same total size)

```{r}
pop_young <- prob_infected |>
  mutate(
    n_people = case_when(
      Age <= 10 ~ 20000,
      Age <= 50 ~ 10000,
      TRUE ~ 0
    ),
    n_infected = n_people * prob_infection
  )

pop_young |>
  summarise(total_infected = sum(n_infected))
```